Making Sequential Programs Deterministic

### 2019-03-25 design.tex
more typesetting
map validity rules

Subtyping rule:

$\infer{\Gamma \vdash x : \tau_2}{\Gamma \vdash x : \tau_1 & \tau_1 \sqsubseteq \tau_2}$

Typing rule for method calls:
\todo{typesetting}

$\infer{\mathit{method}(\overline{a_i}) : \tau_r}{\tau_r \  \mathit{method}(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$

@@

%\todo{It is very important to express the whole section as formal type
%  jugdments, or we will get laughed
%  out of the room as amateurs, during the program committee meeting.  Also,
%  the formalism will help make the ideas precise.
%
%  Type well-formedness rules:\todo{format nicely in \LaTeX}
%
%  antecedent: $\Gamma : \kappa_e <: \kappa_c$
%
%  consequent: $\Gamma : \kappa_c \ \codeid{List}\langle\kappa_e\ 
%  \codeid{Object}\rangle$
%  }
Type validity rules:

$\infer[\sc{ANY TYPE}]{\vdash : \kappa \  \beta}{\kappa : \mathit{Det / NonDet}}$

%$\infer[COLLECTION TYPE]{\vdash : \kappa \  \beta}{\kappa : OrderNonDet & \beta : Collection/Map/Iterator/array}$

$\infer[\sc{ORDERNONDET \ COLLECTION}]{\vdash : \kappa_c \langle\kappa_e \ \beta_e\rangle \beta_c}{\kappa_c : \mathit{OrderNonDet/Det} & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \mathit{Collection/Map/Iterator}}$

$\infer[\sc{ORDERNONDET \ ARRAY}]{\vdash : \kappa_c [\kappa_e \ \beta_e] \beta_c}{\kappa_c : \mathit{OrderNonDet/Det} & \kappa_e \sqsubseteq \kappa_c }$

$\infer[\sc{NONDET \ COLLECTION}]{\vdash : \kappa_c \langle\kappa_e \ \beta_e\rangle \beta_c}{\kappa_c : \mathit{NonDet} & \kappa_e : \mathit{NonDet} & \beta_c \sqsubseteq \mathit{Collection/Map/Iterator}}$

$\infer[\sc{NONDET \ ARRAY}]{\vdash : \kappa_c [\kappa_e \ \beta_e] \beta_c}{\kappa_c : \mathit{NonDet} & \kappa_e : \mathit{NonDet} }$

@@

\subsection{CLIMB-to-top and collection, array locals}\label{climb-rules}

Every type system in the checker framework must define a default type annotation. This annotation will be applied
to every location that isn't explicitly annotated. In our determinism checker, we chose \codeid{@Det} as the default annotation.
The reason for this design choice is that we expect the program to be 
deterministic unless specified by programmer or the program calls nondeterministic library methods that we have annotated 
(Random, Collections, etc).
The framework applies the CLIMB-to-top rule. This rule states that the \textit{top} qualifier 
in the hierarchy is the default for the CLIMB locations: Casts, Locals, Instanceof, and (some) iMplicit Bounds. The rationale
for this rule is as follows:

\begin{itemize}
	\item Local variables are defaulted to top because type refinement is applied to local variables. If a local variable starts as the 
	\textit{top} type, then the Checker Framework refines it to the best (most specific) possible type based on assignments to it. As a 
	result, a programmer rarely writes an explicit annotation on any of those locations.
	Variables defaulted to top include \textit{local variables}, resource variables in the \textit{try-with-resources} construct, variables 
	in \textit{for} statements, and \textit{catch} arguments (known as exception parameters in the Java Language Specification). 
	\textit{Exception parameters} need to have the \textit{top} type because exceptions of arbitrary qualified types can be thrown 
	and the Checker Framework does not provide runtime checks.
	\item \textit{Cast} and \textit{instanceof} types are given the same type as their argument expression. This has the same effect as 
	if they were given the \textit{top} type and then flow-sensitively refined to the type of their argument.
	\item Implicit upper bounds are defaulted to top to allow them to be instantiated in any way. If a user declared class \codeid{C<T> 
		{ ... }}, then the Checker Framework assumes that the user intended to allow any instantiation of the class, and the declaration
	is interpreted as class \codeid{C<T extends @NonDet Object> { ... }} rather than as class \codeid{C<T extends @Det Object> { ... }}. The latter would forbid instantiations such as \codeid{C<@Det String>}, or would require rewriting of code. On the other hand, if a user writes 
	an explicit bound such as class \codeid{C<T extends D> { ... }}, then the user intends some restriction on instantiation and can write a 
	qualifier on the upper bound as desired.
	This rule means that the upper bound of class \codeid{C<T>} is defaulted differently than the upper bound of class \codeid{C<T extends Object>}. 
	This may seem confusing, but it is the least bad option. The more confusing alternative would be for \codeid{Object} to be defaulted 
	differently in class \codeid{C<T extends Object>} and in an instantiation \codeid{C<Object>}, and for the upper bounds to be defaulted differently 
	in class \codeid{C<T extends Object>} and class \codeid{C<T extends Date>}.
	\item \textit{Implicit lower bounds} are defaulted to the \textit{bottom} type, again to allow maximal instantiation. Note that Java does not allow a programmer to express both the upper and lower bounds of a type, but the Checker Framework allows the programmer to specify either or both.
\end{itemize}
With the above rules, the checker framework will automatically annotate a local collection with type \<@NonDet> and its
type arguments will get the type \<@Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
as \codeid{@NonDet List<@Det String>}. Recall from~\ref{collection-rules} that the determinism checker does not allow this type on
any collection. While this is still sound, it will result in the determinism checker reporting a lot of errors because every locally declared 
collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
and annotate their type parameters also with \<@NonDet>. So a local list of Strings gets the default annotation of \codeid{@NonDet List<@NonDet String>}. While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
type, it would could in a high number of false positives. In the following code snippet,
\begin{verbatim}
void test(@Det List<@Det String> argList) {
            List<String> localList = argList;
}    
\end{verbatim}
the checker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
\codeid{@NonDet List<@NonDet> String} and since collections types are invariant.
We therefore recommend that programmers explicitly annotate
all local variables that are collections to reduce false positives. The same reasoning and recommendation applies to arrays.

@@

\subsection{Determinism checker specific rules for precision}

@@

\subsection{Maps discussion}

@@

\subsection{Annotating more classes/providing more specifications}
\begin{itemize}
    \item Type system developer has to annotate methods in the JDK or any other library.
    \item Note: Another source of Nondeterminism we haven't analyzed: File system.
    \item Users can also provide stubs instead. (Steal some text from manual here)
\end{itemize}

>>>>

\begin{figure}
    $\infer[\sc{\mathit{SUBTYPING}}]{\Gamma \vdash x : \tau_2}{\Gamma \vdash x : \tau_1 & \tau_1 \sqsubseteq \tau_2}$

    \bigskip

    $\infer[\sc{\mathit{METHOD \ CALLS}}]{\mathit{method}(\overline{a_i}) : \tau_r}{\tau_r \  \mathit{method}(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$

    \caption{Type checking rules}
    \label{typecheck-rules}
\end{figure}

@@

%\todo{It is very important to express the whole section as formal type
%  jugdments, or we will get laughed
%  out of the room as amateurs, during the program committee meeting.  Also,
%  the formalism will help make the ideas precise.
%
%  Type well-formedness rules:\todo{format nicely in \LaTeX}
%
%  antecedent: $\Gamma : \kappa_e <: \kappa_c$
%
%  consequent: $\Gamma : \kappa_c \ \codeid{List}\langle\kappa_e\ 
%  \codeid{Object}\rangle$
%  }
\begin{figure}
    $\infer[\sc{\mathit{ANY\ TYPE}}]{\vdash : \kappa \  \beta}{\kappa : \mathit{Det / NonDet}}$

    \bigskip

    $\infer[\sc{\mathit{ORDERNONDET \ COLLECTION}}]{\vdash : \kappa_c \  \beta_c \langle\kappa_e \ \beta_e\rangle}{\kappa_c : \mathit{OrderNonDet/Det} & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \mathit{Collection/Iterator}}$

    \bigskip

    $\infer[\sc{\mathit{NONDET \ COLLECTION}}]{\vdash : \mathit{NonDet} \ \beta_c\langle\mathit{NonDet} \ \beta_e\rangle}{\beta_c \sqsubseteq \mathit{Collection/Iterator}}$

    \bigskip

     $\infer[\sc{\mathit{ORDERNONDET \ ARRAY}}]{\vdash : \kappa_e \ \beta_e \ \  \kappa_a []}{\kappa_a : \mathit{OrderNonDet/Det} & \kappa_e \sqsubseteq \kappa_a }$

    \bigskip

    $\infer[\sc{\mathit{NONDET \ ARRAY}}]{\vdash : \mathit{NonDet} \ \beta_e \ \  \mathit{NonDet} []}{}$

    \bigskip

    $\infer[\sc{\mathit{ORDERNONDET \ MAP}}]{\vdash : \kappa_m \  \beta_m \langle\kappa_k \ \beta_k, \kappa_v \ \beta_v\rangle}{\kappa_m: \mathit{OrderNonDet/Det} & \kappa_k \sqsubseteq \kappa_m & \kappa_v \sqsubseteq \kappa_m & \beta_m \sqsubseteq \mathit{Map}}$

    \bigskip

    $\infer[\sc{\mathit{NONDET \ MAP}}]{\vdash : \mathit{NonDet} \ \beta_m\langle\mathit{NonDet} \ \beta_k, \mathit{NonDet} \ \beta_v\rangle}{\beta_m \sqsubseteq \mathit{Map}}$

    \caption{Type well formedness}
    \label{type-validity}
\end{figure}

@@

@@

%\subsection{CLIMB-to-top and collection, array locals}\label{climb-rules}
%
%Every type system in the checker framework must define a default type annotation. This annotation will be applied
%to every location that isn't explicitly annotated. In our determinism checker, we chose \codeid{@Det} as the default annotation.
%The reason for this design choice is that we expect the program to be 
%deterministic unless specified by programmer or the program calls nondeterministic library methods that we have annotated 
%(Random, Collections, etc).
%The framework applies the CLIMB-to-top rule. This rule states that the \textit{top} qualifier 
%in the hierarchy is the default for the CLIMB locations: Casts, Locals, Instanceof, and (some) iMplicit Bounds. The rationale
%for this rule is as follows:
%
%\begin{itemize}
%    \item Local variables are defaulted to top because type refinement is applied to local variables. If a local variable starts as the 
%    \textit{top} type, then the Checker Framework refines it to the best (most specific) possible type based on assignments to it. As a 
%    result, a programmer rarely writes an explicit annotation on any of those locations.
%    Variables defaulted to top include \textit{local variables}, resource variables in the \textit{try-with-resources} construct, variables 
%    in \textit{for} statements, and \textit{catch} arguments (known as exception parameters in the Java Language Specification). 
%    \textit{Exception parameters} need to have the \textit{top} type because exceptions of arbitrary qualified types can be thrown 
%    and the Checker Framework does not provide runtime checks.
%    \item \textit{Cast} and \textit{instanceof} types are given the same type as their argument expression. This has the same effect as 
%    if they were given the \textit{top} type and then flow-sensitively refined to the type of their argument.
%    \item Implicit upper bounds are defaulted to top to allow them to be instantiated in any way. If a user declared class \codeid{C<T> 
%        { ... }}, then the Checker Framework assumes that the user intended to allow any instantiation of the class, and the declaration
%    is interpreted as class \codeid{C<T extends @NonDet Object> { ... }} rather than as class \codeid{C<T extends @Det Object> { ... }}. The latter would forbid instantiations such as \codeid{C<@Det String>}, or would require rewriting of code. On the other hand, if a user writes 
%    an explicit bound such as class \codeid{C<T extends D> { ... }}, then the user intends some restriction on instantiation and can write a 
%    qualifier on the upper bound as desired.
%    This rule means that the upper bound of class \codeid{C<T>} is defaulted differently than the upper bound of class \codeid{C<T extends Object>}. 
%    This may seem confusing, but it is the least bad option. The more confusing alternative would be for \codeid{Object} to be defaulted 
%    differently in class \codeid{C<T extends Object>} and in an instantiation \codeid{C<Object>}, and for the upper bounds to be defaulted differently 
%    in class \codeid{C<T extends Object>} and class \codeid{C<T extends Date>}.
%    \item \textit{Implicit lower bounds} are defaulted to the \textit{bottom} type, again to allow maximal instantiation. Note that Java does not allow a programmer to express both the upper and lower bounds of a type, but the Checker Framework allows the programmer to specify either or both.
%\end{itemize}
%With the above rules, the checker framework will automatically annotate a local collection with type \<@NonDet> and its
%type arguments will get the type \<@Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
%as \codeid{@NonDet List<@Det String>}. Recall from~\ref{collection-rules} that the determinism checker does not allow this type on
%any collection. While this is still sound, it will result in the determinism checker reporting a lot of errors because every locally declared 
%collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
%and annotate their type parameters also with \<@NonDet>. So a local list of Strings gets the default annotation of \codeid{@NonDet List<@NonDet String>}. While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
%type, it would could in a high number of false positives. In the following code snippet,
%\begin{verbatim}
%void test(@Det List<@Det String> argList) {
%List<String> localList = argList;
%}    
%\end{verbatim}
%the checker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
%\codeid{@NonDet List<@NonDet> String} and since collections types are invariant.
%We therefore recommend that programmers explicitly annotate
%all local variables that are collections to reduce false positives. The same reasoning and recommendation applies to arrays.

\subsection{Determinism checker specific rules for precision}\label{precision}

@@

\subsection{Maps discussion}\label{maps}

@@

\subsection{Annotating more classes/providing more specifications}\label{annotating-jdk}
\begin{itemize}
    \item Type system developer has to annotate methods in the JDK or any other library.
    \item Note: Another source of Nondeterminism we haven't analyzed: File system.
    \item Users can also provide stubs instead. (Steal some text from manual here)
\end{itemize}
